---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions, echo = FALSE}
library(knitr)
# always echo.
opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r dataload, cache=TRUE}
if (!file.exists("activity.csv")){
        unzip("activity.zip")
}
activity <- read.csv(file = "activity.csv", colClasses=c(NA,"Date",NA), stringsAsFactors=T, header = TRUE)

```


## What is mean total number of steps taken per day?

###Total number and mean of steps taken per day
```{r totalstepsperday}
library(plyr)

activityPerDay <- ddply(.data = activity, .(date), summarize, stepsPerDay=sum(steps,na.rm = TRUE))
```

###Make a histogram of the total number of steps taken each day

The difference between a histogram and a barplot was researched, based my graph after readying the following pages:

* [Histograms vs Bar Graphs](http://www.shodor.org/interactivate/discussions/HistogramsVsBarGraph/)
* [Plotting distributions (ggplot2)](http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/) 
* [GGPlot Histogram](http://docs.ggplot2.org/current/geom_histogram.html)

```{r histogram_steps_per_day, fig.width = 12, fig.height = 4}

library(ggplot2)
g <- ggplot(activityPerDay) 
g <- g + geom_histogram(binwidth=.5, position="identity", aes(x=date, y=stepsPerDay), stat="identity")
g <- g + labs("Total number of steps taken each day", "Date", "Total steps per day")
g <- g + scale_x_date(breaks = activityPerDay$date)
g <- g + theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

histogramStepsPerDay <- g
# show
histogramStepsPerDay
```


###Calculate and report the mean and median of the total number of steps taken per day
```{r, fig.width = 12, fig.height = 4}
library(plyr)
activityPerDay <- ddply(.data = activity, .(date), summarize, stepsPerDay=sum(steps,na.rm = TRUE), mean= ifelse(is.na(mean(steps, na.rm = T)),0,round(mean(steps, na.rm=T),2)) , median= ifelse(is.na(median(steps, na.rm = T)),0,median(steps, na.rm=T)) )
kable(activityPerDay, col.names = c("Date", "Total steps per day", "Mean", "Median"))
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r timeseriesplot, fig.width = 12, fig.height = 4}
averageByInterval <- ddply(.data = activity, .(interval), summarize, mean=mean(steps, na.rm=T ))

with(averageByInterval,{
        plot(mean ~ interval, type="l")
        abline(v=averageByInterval[averageByInterval$mean==max(averageByInterval$mean),"interval"], col="purple")
        legend(x="topright", cex=0.75, c("Interval with max # of steps"), col=c("purple"), lty=c(1), bty="n")
})
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
averageByInterval[averageByInterval$mean==max(averageByInterval$mean),]
```


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r totalnumbermissingvals}
summary(activity$steps)["NA's"]
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Given that I calculated the mean values in `activityPerDay`, I will use those to replace `NA`, by performing the equivalent of a relational _left join_, and then projecting the right values into the columns. 

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r fixedActivityData}
merged <- merge(activity, activityPerDay, by="date", all.x=T)
fixed0 <- merged[,c("date","interval","mean","steps")]
fixed <- ddply(fixed0, .(date, interval), function(r){ c(ifelse(is.na(r$steps),r$mean,r$steps))})
colnames(fixed) <- c("date","interval","steps")
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r histogram_steps_per_day_fixed, fig.width = 12, fig.height = 4}

library(ggplot2)
activityPerDayFixed <- ddply(.data = activity, .(date), summarize, stepsPerDay=sum(steps,na.rm = TRUE))
g <- ggplot(activityPerDayFixed) 
g <- g + geom_histogram(binwidth=.5, position="identity", aes(x=date, y=stepsPerDay), stat="identity")
g <- g + labs("Total number of steps taken each day", "Date", "Total steps per day")
g <- g + scale_x_date(breaks = activityPerDayFixed$date)
g <- g + theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))
# show

histogramStepsPerDayFixed <- g

g
```

4. a) Do these values differ from the estimates from the first part of the assignment? 

They do not seem to differ.

```{r}
# based on http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/#put-two-potentially-unrelated-plots-side-by-side-pushviewport
library(grid)
pushViewport(viewport(layout = grid.layout(2, 1)))
print(histogramStepsPerDay, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(histogramStepsPerDayFixed, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
```

4. b) What is the impact of imputing missing data on the estimates of the total daily number of steps?

Given my initial load/cleanup of data (not taking NAs in account), the first histogram and the histogram resulting from inputing the data do not differ.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
library(stringr)

activity <- transform(activity, typeOfDay=ifelse(str_detect(weekdays(date),"Sunday|Saturday"),"weekend","weekday"))

```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r totalstepsperdayByWeekdays}
library(plyr)
library(lattice)

averageByIntervalWeekdays <- ddply(.data = activity, .(interval, date), summarize, mean=mean(steps, na.rm=T ), typeOfDay=typeOfDay)

par(mfrow = c(1,2))
xyplot(mean ~ interval | typeOfDay, averageByIntervalWeekdays, type="l", xlab = "Interval", ylab="Number of Steps")

```

