---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions, echo = FALSE}
library(knitr)
# always echo.
opts_chunk$set(echo = TRUE, fig.width = 12, fig.height = 5, dev='png')
```

## Loading and preprocessing the data
```{r dataload, cache=TRUE}
if (!file.exists("activity.csv")){
        unzip("activity.zip")
}
activity <- read.csv(file = "activity.csv", colClasses=c(NA,"Date",NA), stringsAsFactors=T, header = TRUE)

```


## What is mean total number of steps taken per day?

Total number and mean of steps taken per day is calculated by grouping all rows by date, then summarizing the steps count for all intervals in that day. Results are saved in a new dataframe.

<a id="totalstepsperday"></a>

```{r totalstepsperday}
library(plyr)

activityPerDay <- ddply(.data = activity, .(date), summarize, stepsPerDay=sum(steps,na.rm = TRUE))
```

###Make a histogram of the total number of steps taken each day

I researched the difference between a histogram and a barplot, then based my graphs after reading the following pages:

* [Histograms vs Bar Graphs](http://www.shodor.org/interactivate/discussions/HistogramsVsBarGraph/)
* [Plotting distributions (ggplot2)](http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/) 
* [GGPlot Histogram](http://docs.ggplot2.org/current/geom_histogram.html)

```{r histogram_steps_per_day}

library(ggplot2)
g <- ggplot(activityPerDay) 
g <- g + geom_histogram(colour = "cadetblue", fill = "white",binwidth=.1, position="identity", aes(x=date, y=stepsPerDay), stat="identity")
g <- g + ggtitle("Total number of steps taken each day") + xlab("Date") + ylab("Total steps per day")
g <- g + scale_x_date(breaks = activityPerDay$date)
g <- g + theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

g <- g +  theme(axis.title.x = element_text(color="forestgreen", vjust=-0.35),
      axis.title.y = element_text(color="cadetblue" , vjust=0.35) ) 


# save it for later
histogramStepsPerDay <- g

# show
g
```


###Calculate and report the mean and median of the total number of steps taken per day

Following is a table displaying the total number of steps taken for each day, along with its mean and median.


```{r meanAndMedianOfStepsByDay}
library(plyr)

activityPerDay <- ddply(.data = activity, .(date), summarize, stepsPerDay=sum(steps,na.rm = TRUE), mean= ifelse(is.na(mean(steps, na.rm = T)),0,round(mean(steps, na.rm=T),2)) , median= ifelse(is.na(median(steps, na.rm = T)),0,median(steps, na.rm=T)) )

#kable(activityPerDay, col.names = c("Date", "Total steps per day", "Mean", "Median"), caption="Mean and median of the total number of steps taken per day")
activityPerDay
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r timeseriesplot}
averageByInterval <- ddply(.data = activity, .(interval), summarize, mean=mean(steps, na.rm=T ))

highestInterval <- averageByInterval[averageByInterval$mean==max(averageByInterval$mean),"interval"]
highestIntervalLabel <- paste0("Interval with max # of steps (",highestInterval,")")

with(averageByInterval,{
        plot(mean ~ interval, type="l", ylab = "Average number of steps", xlab="Interval")
        abline(v=highestInterval, col="purple", lwd=2)
        title(main="Average number of steps taken, averaged across all days, per 5 min interval")
        legend(x="topright", 
                cex=0.75, c(highestIntervalLabel), 
                col=c("purple"), lty=c(1), bty="n", lwd = c(2))

})
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r}
averageByInterval[averageByInterval$mean==max(averageByInterval$mean),]
```


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r totalnumbermissingvals}
summary(activity$steps)["NA's"]
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Given that I calculated the mean values in `activityPerDay` [above](#totalstepsperday), I will use those mean values to replace the original `NA`s by performing the equivalent of a relational _left join_, and then projecting the right values, then renaming the resulting columns to match the original ones. 

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r fixedActivityData}
merged <- merge(activity, activityPerDay, by="date", all.x=T)
fixed0 <- merged[,c("date","interval","mean","steps")]
fixed <- ddply(fixed0, .(date, interval), function(r){ c(ifelse(is.na(r$steps),r$mean,r$steps))})
colnames(fixed) <- c("date","interval","steps")
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r histogram_steps_per_day_fixed}

library(ggplot2)
activityPerDayFixed <- ddply(.data = activity, .(date), summarize, stepsPerDay=sum(steps,na.rm = TRUE))

g <- ggplot(activityPerDayFixed) 
g <- g + geom_histogram(colour = "chocolate", fill = "white",binwidth=.1, position="identity", aes(x=date, y=stepsPerDay), stat="identity")
g <- g + ggtitle("Total number of steps taken each day") + xlab("Date") + ylab("Total steps per day")
g <- g + scale_x_date(breaks = activityPerDayFixed$date)
g <- g + theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

g <- g +  theme(axis.title.x = element_text(color="coral4", vjust=-0.35),
      axis.title.y = element_text(color="chocolate" , vjust=0.35) ) 


histogramStepsPerDayFixed <- g

g
```

4. a) Do these values differ from the estimates from the first part of the assignment? 

Visual inspection of both histograms suggests they *do not differ*.

```{r}
# based on http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/#put-two-potentially-unrelated-plots-side-by-side-pushviewport
library(grid)
pushViewport(viewport(layout = grid.layout(2, 1)))
print(histogramStepsPerDay, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(histogramStepsPerDayFixed, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
```

4. b) What is the impact of imputing missing data on the estimates of the total daily number of steps?

Given my initial load/cleanup of data (not taking NAs in account), the first histogram and the histogram resulting from inputing the data do not differ.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
library(stringr)

activity <- transform(activity, typeOfDay=ifelse(str_detect(weekdays(date),"Sunday|Saturday"),"weekend","weekday"))

# show random sample to demonstrate it worked
set.seed(10)
activity[sample(nrow(activity),10),]
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r totalstepsperdayByWeekdays}
library(plyr)
library(lattice)

averageByIntervalWeekdays <- ddply(.data = activity, .(interval, date), summarize, mean=mean(steps, na.rm=T ), typeOfDay=typeOfDay)

xyplot(mean ~ interval | typeOfDay, averageByIntervalWeekdays, type="l", xlab = "Interval", ylab="Number of Steps", layout = c(1,2))
```

